\chapter{Conclusions and Future Work}
\label{chap:conclusions}

\par This thesis designed, implemented, and evaluated a real-time intrusion detection system for IoT networks built on the CICIoT2023 dataset, delivering a reproducible offline training pipeline and an interactive web demonstration. The data pipeline ingests the raw CICIoT2023 captures, removes the large exact-duplicate fraction before sampling, subsamples each class to a 200{,}000-row cap, and writes a compressed columnar cache, reproducing byte-identical splits from the same seed and source files. A feature analysis characterises the 39 public features through variance and inter-feature correlation, pruning 14 near-zero-variance indicators and two perfectly collinear continuous pairs to leave a 25-feature input. On that input a multi-layer perceptron was trained at binary and 8-class attack-family granularities on a stratified random split and compared against a Random Forest baseline under identical metrics, so the comparison is controlled. The MLP's softmax confidences were calibrated by temperature scaling at no cost to predicted labels. The NEURAL IDS web demonstration pairs a Gradio/FastAPI backend with a React dashboard that accepts pre-extracted CSVs or raw packet captures, classifies flows in under 10 ms, and shows confidence scores, per-class probabilities, and a session analysis history, deployed on Hugging Face Spaces for remote access.

\par The experiments yielded four main findings. Both models generalise from train to test, with an accuracy drop below 8 percentage points, though on a random split this small gap partly reflects within-session redundancy rather than genuine generalisation. The binary weighted F1 (around 0.90 for the MLP) clears the 0.85 bar set by NFR-2, while the 8-class figure (around 0.73 for the MLP and 0.77 for the Random Forest) sits well below it, which is the expected price of fine-grained family discrimination when the features describe only transport-layer behaviour. At the family level, Mirai is the easiest class, reaching near-perfect recall (0.99) thanks to its distinctive high-rate, fixed-payload traffic, whereas DDoS and DoS are frequently confused with each other (up to 36\% cross-confusion) because both are volumetric floods separated mainly by the number of coordinated source hosts, a distinction the fixed-window feature representation does not capture reliably. The hardest classes are the application-layer attacks: Spoofing and Brute Force are weakest at 0.46 recall, Reconnaissance reaches 0.51, and the web family 0.52, with a large fraction of Spoofing flows classified as Benign; these families are distinguished mainly by payload content, which the header-and-metadata features do not capture, so the ceiling here is set by the data rather than the classifier. Finally, the Random Forest outperforms the MLP by roughly 4 to 5 percentage points on test accuracy and weighted F1, consistent with the tabular-learning literature on heterogeneous, non-spatial features at sample sizes in the low millions; the MLP is retained as the serving model because its inference is a single portable PyTorch call and the gap is small enough to keep the system within its targets.

\par Several limitations bound the scope of these conclusions. The 25 retained features come from packet headers and metadata only, with no application-layer content (HTTP bodies, DNS query strings, TLS certificate fields), which caps detection accuracy for application-layer attacks and means the results do not generalise to settings where the attack signal lives in the payload. The captures come from a controlled testbed of 105 real IoT devices, so the attacker-victim topology, the absence of background noise, and the isolated execution of individual attack types do not reflect multi-stage campaigns, lateral movement, or the traffic diversity of a production network. Only the random split is reported here; because it does not keep each capture wholly on one side of the train/test boundary, near-duplicate flows leak across it, so the numbers are an upper bound on deployment performance, and leakage-aware splits would give a more conservative measure. The work uses the 39-feature public CSV release, and the seven features missing from it (source rate, destination rate, URG count, Magnitude, Radius, Covariance, Weight) may carry discriminative information, particularly for the weaker families, so the numbers here are not directly comparable to studies using the full 46-feature variant. Temperature scaling reduces but does not eliminate calibration error (8-class ECE 0.089 after scaling): a single global temperature cannot correct per-class miscalibration, and calibration was measured in-distribution, so behaviour on genuinely out-of-distribution traffic, where softmax outputs are known to be overconfident \cite{Guo2017}, remains uncharacterised.

\par These findings and limitations suggest several directions for future work. The most immediate is to evaluate on leakage-aware splits, namely a group-wise split that keeps each capture whole and a temporal split that trains on earlier captures and tests on later ones, which would quantify how much of the random-split performance is inflated by within-session redundancy. A second is to rerun the pipeline on the full 46-feature variant once it is available from a verified source, allowing a direct comparison with published baselines and possibly improving detection of the weaker families. A third is to move beyond the global temperature scaling already applied towards per-class scaling and per-class decision thresholds, so the detection-rate and false-alarm trade-off can be tuned per attack family (for instance a lower threshold for the low-recall Spoofing class) rather than fixed at the argmax. A fourth is to extend the feature set with TLS-visible metadata such as certificate fields, SNI, and session-length distributions, which would reach partially encrypted traffic and may help the web-based attacks that the transport-layer representation currently limits. Beyond these, deploying the serving component on a real IoT gateway with a periodic retraining schedule driven by misclassified flows would test the pipeline's generalisation outside the laboratory and address the static, offline nature of the present model.
