%\documentclass[12pt]{scrreprt}
\documentclass[12pt]{report} 

% language may be romanian or english (default is english)
% type may be bachelor or master (default is bachelor)
\usepackage[language=english, type=bachelor]{style}

%\geometry{a4paper,top=2.5cm,left=3cm,right=2.5cm,bottom=2.5cm}
%in style
%controlling the appearance of your headers and footers
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{microtype}
\usepackage{tikz}
\usetikzlibrary{arrows, positioning, shapes.geometric, fit, backgrounds}
\usepackage{csquotes}
\usepackage{fancyhdr}
\pagestyle{fancy}
\setlength{\headheight}{14.5pt}
\lhead{}
\chead{}
\renewcommand{\headrulewidth}{0.2pt}
\renewcommand{\footrulewidth}{0.2pt}

% Allow LaTeX more flexibility breaking long paragraphs to avoid overfull hboxes.
\emergencystretch=2em

% Tighten the vertical space around floats and captions (body line spacing is unchanged).
\setlength{\textfloatsep}{8pt plus 2pt minus 2pt}
\setlength{\floatsep}{8pt plus 2pt minus 2pt}
\setlength{\intextsep}{8pt plus 2pt minus 2pt}
\setlength{\abovecaptionskip}{4pt}
\setlength{\belowcaptionskip}{2pt}

\begin{document}

\specialization{Artificial Intelligence}	
\title{Real-Time Intrusion Detection for Web Applications Using Deep Learning}					   
\author{Rares Marta}											
\supervisor{Alexandra--Ioana Albu}				
				
\maketitle


\pagenumbering{roman}

\cleardoublepage
ABSTRACT
\vspace{0.5cm}	
\hrule
\vspace{0.5cm}	

This thesis designs, implements, and evaluates a real-time intrusion detection system (IDS) for IoT networks using deep learning on the CICIoT2023 benchmark dataset, delivering a complete, reproducible pipeline from raw packet captures to live per-flow classification.

The system is built around a two-hidden-layer Multi-Layer Perceptron (MLP) trained at binary and 8-class (attack-family) granularities on a stratified random train/validation/test split, following common practice on CICIoT2023. The MLP is compared against a Random Forest baseline under identical conditions. On the binary task both models exceed the 0.85 weighted-F1 target, with the Random Forest reaching $\approx$0.92 and the MLP $\approx$0.90; on the harder 8-class task the Random Forest leads at $\approx$0.77 and the MLP attains $\approx$0.73, neither reaching 0.85, a shortfall traced to the limits of the public transport-layer features for separating application-layer attack families. Feature analysis identifies the per-flow packet count, the HTTPS/HTTP indicators, and header length as the most discriminative features, and prunes 14 of the 39 near-zero-variance CICIoT2023 features to leave a 25-feature model. Temperature scaling calibrates the MLP's softmax confidences.

The practical component is a web demonstration, NEURAL IDS, pairing a FastAPI backend with a React dashboard that accepts pre-extracted feature CSVs or raw packet captures, classifies flows in under 10 milliseconds, and displays confidence scores and per-class probability breakdowns. It correctly flags a synthetic SYN-flood capture as an attack (every flow flagged as Attack, mean calibrated confidence 99.6\%) and classifies benign browsing as Benign (98\% of flows), validating the core classification functionality required by the project specification.

% Show only chapters and sections in the TOC (subsections inflate it to 3 pages).
\setcounter{tocdepth}{1}
{\setstretch{1.0}\tableofcontents}


\newpage
\pagenumbering{arabic}

\input{chapters/chapter1_introduction}
%\addcontentsline{toc}{chapter}{Introduction}

\input{chapters/chapter2}
\input{chapters/chapter3_ml_methodology}
\input{chapters/chapter4_results}
\input{chapters/chapter5_requirements}
\input{chapters/chapter6_application}

\input{chapters/chapter7_conclusions}
%\addcontentsline{toc}{chapter}{Conclusions}

\par % close the final body paragraph under 1.5 spacing before the single-spaced bibliography group
\begingroup
\setstretch{1.0}
\small
\bibliographystyle{plain}
\bibliography{references}
\endgroup

\end{document}